Separation in Generic Extensions for Isabelle

### 2019-01-05 LICS/hacking.tex
eds varias
ediciones varias, most sugeridas por Miguel

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%          
\section{Hacking of \isatt{ZF-Constructible}}
\label{sec:hack-constructible}
In \cite{paulson_2003}, Paulson presented his formalization of the
relative consistency of the Axiom of Choice. This development is
included inside the Isabelle distribution with session name
\isatt{ZF-Constructible}. The main technical devices, invented by
G\"odel for this purpose, are \emph{relativization} and
\emph{absoluteness}. In a nutshell, to relativize a formula $\phi$ to
a class $C$, it is enough to restrict its quantifiers to $C$. The
example of \isatt{upair\_ax} in
Section~\ref{sec:axioms-models-set-theory}, the relativized version of
the Pairing Axiom, is extracted from \texttt{Relative}, one of the
core theories of \isatt{ZF-Constructible}. On the other hand, $\phi$
is \emph{absolute} for $C$ if it is equivalent to its relativization,
meaning that the statement made $\phi$ coincides with what $C$
``believes'' $\phi$ is saying. Paulson shows that under certain
hypotheses  on a class $M$ (condensed in the locale \isatt{M\_trivial}), a plethora of
absoluteness and closure results can be proved about $M$.

The development of forcing, and the study of ctms in general, takes
absoluteness as a starting point. We were not able to work with
\isatt{ZF-Constructible} right out-of-the-box. The main reason is that
we can't expect state the ``class version'' of Replacement for a
\emph{set} $M$ by
using first-order formulas, since predicates \isatt{P::"i=>o"} can't
be proved to be only the definable ones. Therefore, we had to make
some modifications to the various sets of hypotheses in several
locales to make the results available as tools for the present and
future developments.

@@

The most notable changes, located in the theory \texttt{Relative}, are
the following:
\begin{enumerate}
\item\label{item:1} We eliminated the requirement that the relative Axiom of Replacement
  is satisfied by a class $M$ to be in the locale \isatt{M\_trivial}. 
\item\label{item:2} We moved the requirement of the Powerset Axiom to \isatt{M\_basic}. 
\item\label{item:3} We replaced the need that the set of natural numbers is in $M$ by the
  milder hypothesis that $M(0)$. Actually, most results should follow
  by only assuming that $M$ is nonempty.
\end{enumerate}

As a consequence of Item~\ref{item:1}, the lemma
\isatt{strong\_replacementI} is no longer valid and was commented
out.

We moved the requirement $M(\mathtt{nat})$ to the locale
\isatt{M\_trancl} (inside the theory \isatt{WF\_absolute}), where it is needed for the first time. Some results,
for instance \isatt{rtran\_closure\_mem\_iff} and 
\isatt{iterates\_imp\_wfrec\_replacement} had to be moved inside that
locale.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%          
\section{Hacking of \isatt{ZF-Constructible}}
\label{sec:hack-constructible}
In \cite{paulson_2003}, Paulson presented his formalization of the
relative consistency of the Axiom of Choice. This development is
included inside the Isabelle distribution under the session 
\isatt{ZF-Constructible}. The main technical devices, invented by
G\"odel for this purpose, are \emph{relativization} and
\emph{absoluteness}. In a nutshell, to relativize a formula $\phi$ to
a class $C$, it is enough to restrict its quantifiers to $C$. The
example of \isatt{upair\_ax} in
Section~\ref{sec:axioms-models-set-theory}, the relativized version of
the Pairing Axiom, is extracted from \texttt{Relative}, one of the
core theories of \isatt{ZF-Constructible}. On the other hand, $\phi$
is \emph{absolute} for $C$ if it is equivalent to its relativization,
meaning that the statement made $\phi$ coincides with what $C$
``believes'' $\phi$ is saying. Paulson shows that under certain
hypotheses  on a class $M$ (condensed in the locale \isatt{M\_trivial}), a plethora of
absoluteness and closure results can be proved about $M$.

The development of forcing, and the study of ctms in general, takes
absoluteness as a starting point. We were not able to work with
\isatt{ZF-Constructible} right out-of-the-box. The main reason is that
we can't expect to state the ``class version'' of Replacement for a
\emph{set} $M$ by
using first-order formulas, since predicates
\isatt{P::"i\isasymRightarrow o"} can't
be proved to be only the definable ones. Therefore, we had to make
some modifications in several
locales to make the results available as tools for the present and
future developments.

@@

The most notable changes, located in the theories \texttt{Relative}
and \isatt{WF\_absolute}, are
the following:
\begin{enumerate}
\item\label{item:1} The locale \isatt{M\_trivial}
  does not assume that the underlying class $M$ satisfies the relative
  Axiom of replacement.  As a consequence, the lemma
  \isatt{strong\_replacementI} is no longer valid and was commented
  out.
\item\label{item:2}  Originally the Powerset Axiom was assumed by the
  locale \isatt{M\_trivial},   we moved this requirement to \isatt{M\_basic}. 
\item\label{item:3} We replaced the need that the set of natural
  numbers is in $M$ by the   milder hypothesis that $M(0)$. Actually,
  most results should follow 
  by only assuming that $M$ is nonempty.
\item We moved the requirement $M(\mathtt{nat})$ to the locale
  \isatt{M\_trancl}, where it is needed for the first time. Some results,
  for instance \isatt{rtran\_closure\_mem\_iff} and 
  \isatt{iterates\_imp\_wfrec\_replacement} had to be moved inside that
  locale. 
\end{enumerate}
Because of these changes, some theory files from the
\isatt{ZF-Constructible} session have been included among ours.
